---
title: "COVID-19 PBMC Analysis with tinydenseR: Identifying Dysregulated Cell States"
author: "Pedro Milanez-Almeida"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{COVID-19 PBMC Analysis with tinydenseR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  out.width = "100%",
  eval = TRUE
)
```

## Introduction

This vignette demonstrates how `tinydenseR` can be applied to a real-world
clinical scRNA-seq dataset to identify dysregulated cell states in COVID-19
patients. We use PBMC data from the COMBAT (COvid-19 Multi-omics Blood
ATlas) consortium to show how differential density analysis combined with
PLS-based decomposition (plsDE) reveals:

1. **Dysregulated monocyte states** in early critical COVID-19 relative to mild
   disease — captured by plsDE component 1 (plsDE1) scores and loadings.
2. **Expanded immature cell populations** and **depleted pDCs** in critical cases — captured by plsDE component 2 (plsDE2).

### Analysis Overview

The core workflow follows four steps:

1. `RunTDR()` — build landmarks from an on-disk count matrix and compute
   sample-level density profiles.
2. Visualize landmark and sample embeddings — inspect the UMAP of landmarks
   and the sample PCA to confirm biological structure.
3. `get.lm()` with contrasts — perform differential density analysis while
   controlling for covariates (sex, age, time since symptom onset).
4. `get.plsDE()` — decompose the density contrast into interpretable
   components that jointly capture differential abundance and differential
   expression.

### Data Source

The COMBAT dataset is available from CELLxGENE
([link](https://datasets.cellxgene.cziscience.com/687c09ff-731a-4e3d-ac07-4c29c33a6338.h5ad)). The original study is described in:

> COMBAT Consortium. *A blood atlas of COVID-19 defines hallmarks of disease
> severity and specificity.* Cell 185, 916–938 (2022).

## Setup

### Load Packages

```{r load-packages}
library(tinydenseR)
library(dplyr)
library(forcats)
library(ggplot2)
library(patchwork)
library(limma)
library(anndataR)
library(BPCells)
```

### Load and Prepare Data

We read the COMBAT `.h5ad` file, keeping data on disk via `anndataR` and
`BPCells` for memory efficiency. This section shows the key preprocessing
steps; adjust file paths to match your local setup.

```{r load-data}
# Path to the downloaded h5ad file
file_path <- "~/sandbox/COVID19_COMBAT/derived_data/COMBAT.h5ad"

# Access h5ad keeping data on disk
h5ad <- 
  anndataR::read_h5ad(
    path = file_path,
    as = "HDF5AnnData")

# Extract metadata and feature information
feature_metadata <- h5ad$var
metadata <- h5ad$obs
```

### Quality Control and Filtering

We apply basic QC thresholds and restrict the analysis to healthy volunteers
(HV) and COVID-19 patients with mild, severe, or critical disease. Samples
with fewer than 200 cells are excluded.

```{r qc-filter}
# Basic QC: minimum 1000 UMI and 500 genes per cell
keep.QC.cells <- 
  (metadata$QC_total_UMI >= 1000) & 
  (metadata$QC_ngenes >= 500)

# Keep only HV, MILD, SEV, and CRIT patients with >= 200 cells
smpl.to.keep <-
  metadata$scRNASeq_sample_ID[
    metadata$Source %in% c("HV", "COVID_MILD", "COVID_SEV", "COVID_CRIT")
  ] |> 
  table() |>
  (\(x) names(x[x >= 200]))()

metadata <-
  metadata[keep.QC.cells & (metadata$scRNASeq_sample_ID %in% smpl.to.keep), ]
```

### Prepare the On-Disk Count Matrix

We use `BPCells` to store the count matrix on disk and subset it to QC-passing
cells with unique gene symbols.

```{r ondisk-matrix}
# Remove duplicated gene symbols
keep.unique.genes <- !duplicated(feature_metadata$feature_name)
feature_metadata <- feature_metadata[keep.unique.genes, ] |> droplevels()

# Write BPCells on-disk matrix
ondisk_path <- 
  "~/sandbox/COVID19_COMBAT/derived_data/COMBAT_COVID_ondisk/"

ondisk.data <-
  BPCells::open_matrix_anndata_hdf5(
    path = file_path,
    group = "/X")

BPCells::write_matrix_dir(
  mat = ondisk.data, 
  dir = ondisk_path,
  overwrite = TRUE)

# Re-open and subset
ondisk.mat <-
  BPCells::open_matrix_dir(dir = ondisk_path)

ondisk.mat <-
  ondisk.mat[keep.unique.genes, keep.QC.cells & (metadata$scRNASeq_sample_ID %in% smpl.to.keep)]

rownames(x = ondisk.mat) <- feature_metadata$feature_name
```

### Prepare Metadata Covariates

We encode disease severity as ordered factor levels and binarize time since
symptom onset into "early" and "late" to allow for non-linear effects. We
also create a combined `Source_TSO.binary` variable for the contrast design.

```{r prepare-metadata}
metadata <-
  metadata |>
  dplyr::mutate(
    Source = factor(
      x = Source,
      levels = c("HV", "COVID_MILD", "COVID_SEV", "COVID_CRIT")),
    TimeSinceOnset = ifelse(
      test = Source != "HV", 
      yes = TimeSinceOnset, 
      no = -1)
  ) |>
  droplevels()
```

## Step 1: Build Landmarks with `RunTDR()`

`RunTDR()` is the main entry point. It creates landmark cells, builds a
neighborhood graph, computes sample-level density profiles, and maps cell type
annotations — all in a single call.

```{r run-tdr}
lm.cells <-
  tinydenseR::RunTDR(
    x = ondisk.mat,
    .sample.var = "scRNASeq_sample_ID", 
    .cell.meta = metadata,
    .assay.type = "RNA",
    .celltype.vec = "cell_type",
    .label.confidence = 0.5,
    .verbose = FALSE
  )
```

## Step 2: Visualize Landmark and Sample Embeddings

Before statistical testing, we inspect the landmark UMAP and sample PCA to
confirm that the data exhibits the expected biological structure.

### Landmark UMAP

The UMAP of landmarks reveals major immune cell populations. Each point is a
landmark colored by its unsupervised cluster assignment.

```{r umap-clustering, fig.width=10, fig.height=4}
tinydenseR::plotUMAP(
  x = lm.cells,
  .plot.title = "COMBAT landmarks"
)
```

We can overlay reference-based cell type annotations to identify clusters:

```{r umap-celltype, fig.width=10, fig.height=4}
tinydenseR::plotUMAP(
  x = lm.cells,
  .feature = lm.cells$landmark.annot$celltyping$ids,
  .plot.title = "COMBAT landmarks (cell types)"
)  +
  ggplot2::guides(color = ggplot2::guide_legend(
    override.aes = list(size = 5),
    ncol = 2))
```

We can also project individual gene expression (e.g., canonical markers) onto
the UMAP to validate cell identities:

```{r umap-markers, fig.width=10, fig.height=4}
# Classical monocyte marker
p1 <- 
  tinydenseR::plotUMAP(
  x = lm.cells,
  .feature = lm.cells$landmarks[, "LYZ"],
  .plot.title = "LYZ (classical monocytes)")
# MHC class II
p2 <- 
  tinydenseR::plotUMAP(
    x = lm.cells,
     .feature = lm.cells$landmarks[, "HLA-DRA"],
     .plot.title = "HLA-DRA (MHC class II)")
# T cell marker
p3 <- 
  tinydenseR::plotUMAP(
    x = lm.cells,
     .feature = lm.cells$landmarks[, "CD3E"],
     .plot.title = "CD3E (T cells)")

p1 | p2 | p3
```

### Sample PCA

The sample PCA provides a low-dimensional view of inter-sample variation based
on their density profiles across landmarks. The color encodes sample library
size (log10 total cells).

```{r sample-pca, fig.width=10, fig.height=4}
tinydenseR::plotSampleEmbedding(
  x = lm.cells,
  .embedding = "pca",
  .color.by = "log10.n.cells"
)
```

## Step 3: Differential Density Analysis with `get.lm()`

### Encoding Covariates

We binarize time since onset into "early" vs. "late" and create a composite
factor that crosses disease severity with timing. This allows us to define
biologically meaningful contrasts while controlling for confounders.

```{r encode-covariates}
# Binarize TimeSinceOnset
lm.cells@metadata <-
  dplyr::mutate(
    .data = lm.cells@metadata,
    TSO.binary = 
      dplyr::case_when(
        Source == "HV" ~ "uninfected",
        Source != "HV" ~ ifelse(
          TimeSinceOnset > median(TimeSinceOnset[Source != "HV"]),
          "late", "early")
      ) |>
      factor(levels = c("uninfected", "early", "late"))
  )

# Standardize Age encoding
lm.cells$metadata$Age <-
  make.names(lm.cells$metadata$Age) |>
  gsub(pattern = "^X", replacement = "Age.", fixed = FALSE)

# Create composite Source x TSO variable
lm.cells$metadata$Source_TSO.binary <-
  paste(lm.cells$metadata$Source,
        lm.cells$metadata$TSO.binary,
        sep = "_") |>
          gsub(pattern = "^HV_|^COVID_", 
          replacement = "",
           fixed = FALSE) |>
            factor(levels = c("uninfected", "MILD_early", "MILD_late", "SEV_early", "SEV_late", "CRIT_early", "CRIT_late"))
```

### Design and Contrast Matrices

We set up a cell-means design with `~ 0 + Source_TSO.binary + sex + Age`,
so that `sex` and `Age` are controlled for in every contrast. The contrast
matrix defines the biologically interesting comparisons.

```{r design-contrast}
# Design matrix (cell-means model controlling for sex and Age)
.design <- 
  model.matrix(~ 0 + Source_TSO.binary + sex + Age,
               data = lm.cells$metadata) |> 
  (\(x) {
    colnames(x) <- gsub("^Source_TSO.binary", "", colnames(x))
    x
  })()

# Contrast matrix
.contrast.matrix <- 
  limma::makeContrasts(
    # Average early COVID vs healthy
    earlyInfectionVSuninfected =
      ((MILD_early + SEV_early + CRIT_early) / 3) -
      uninfected,
    
    # Average late COVID vs early COVID
    lateInfectionVSearlyInfection =
      ((MILD_late + SEV_late + CRIT_late) / 3) -
      ((MILD_early + SEV_early + CRIT_early) / 3),
    
    # Critical vs mild within early timepoint
    CRIT_earlyVSMILD_early =
      CRIT_early - MILD_early,
    
    levels = .design
  )
```

### Fit the Model

`get.lm()` fits a limma model to the log-transformed fuzzy density profiles
across landmarks, applying the contrast matrix. This identifies which
landmarks are differentially abundant between conditions — analogous to
finding differentially expressed genes in bulk RNA-seq, but at the level of
cell density.

```{r fit-lm}
lm.cells <- 
  tinydenseR::get.lm(
    x = lm.cells,
    .design = .design,
    .contrasts = .contrast.matrix,
    .model.name = "contrast.Source_TSO.binary",
    .verbose = TRUE
  )
```

### Visualize Differential Density

We can now overlay the q-values and fold-change directions onto the landmark
UMAP. This highlights which cell states are significantly expanding or
depleting. The key contrast for the main result is early CRIT vs MILD.

```{r da-umap, fig.width=10, fig.height=4}
# get coloring based on fold-change direction and significance for each contrast
q.earlyVsHV.c <-
  ifelse(
    test = lm.cells$results$lm$contrast.Source_TSO.binary$fit$coefficients[, "earlyInfectionVSuninfected"] < 0,
    yes = "lower density",
    no = "higher density") |>
      ifelse(
        test = lm.cells$results$lm$contrast.Source_TSO.binary$fit$pca.weighted.q[, "earlyInfectionVSuninfected"] < 0.1,
        no = "not sig.") |>
          factor(
            levels = c("lower density", "not sig.", "higher density"))

q.lateVsEarly.c <-
  ifelse(
    test = lm.cells$results$lm$contrast.Source_TSO.binary$fit$coefficients[, "lateInfectionVSearlyInfection"] < 0,
    yes = "lower density",
    no = "higher density") |>
      ifelse(
        test = lm.cells$results$lm$contrast.Source_TSO.binary$fit$pca.weighted.q[, "lateInfectionVSearlyInfection"] < 0.1,
        no = "not sig.") |>
          factor(
            levels = c("lower density", "not sig.", "higher density"))

q.critVsMild.c <-
  ifelse(
    test = lm.cells$results$lm$contrast.Source_TSO.binary$fit$coefficients[, "CRIT_earlyVSMILD_early"] < 0,
    yes = "lower density",
    no = "higher density") |>
      ifelse(
        test = lm.cells$results$lm$contrast.Source_TSO.binary$fit$pca.weighted.q[, "CRIT_earlyVSMILD_early"] < 0.1,
        no = "not sig.") |>
          factor(
            levels = c("lower density", "not sig.", "higher density"))

# Early COVID vs healthy
q.earlyVsHV <-
  tinydenseR::plotUMAP(
    x = lm.cells,
    .feature = q.earlyVsHV.c,
    .plot.title = "early COVID vs healthy",
    .cat.feature.color = unname(obj = Color.Palette[1, c(1, 6, 2)]),
    .color.label = "q < 0.1"
  ) +
  ggplot2::theme(plot.subtitle = ggplot2::element_blank())

# Late vs early COVID
q.lateVsEarly <-
  tinydenseR::plotUMAP(
    x = lm.cells,
    .feature = q.lateVsEarly.c,
    .plot.title = "COVID: late vs early",
    .cat.feature.color = unname(obj = Color.Palette[1, c(1, 6, 2)]),
    .color.label = "q < 0.1"
  ) +
  ggplot2::theme(plot.subtitle = ggplot2::element_blank())

# Early CRIT vs MILD — the key contrast
q.critVsMild <-
  tinydenseR::plotUMAP(
    x = lm.cells,
    .feature = q.critVsMild.c,
    .plot.title = "early: CRIT vs MILD",
    .cat.feature.color = unname(obj = Color.Palette[1, c(1, 6, 2)]),
    .color.label = "q < 0.1"
  ) +
  ggplot2::theme(plot.subtitle = ggplot2::element_blank())

(q.earlyVsHV | q.lateVsEarly | q.critVsMild) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "bottom",
                 legend.justification = "center")
```

### Sample Embeddings by Contrast

We can also compute supervised sample embeddings (partial-effect PCs) for
each contrast, providing a low-dimensional score that summarizes how each
sample deviates along a contrast direction.

```{r sample-embed, fig.width=10, fig.height=4}
# Compute sample embeddings for each contrast
for (contrast in colnames(lm.cells$results$lm$contrast.Source_TSO.binary$fit$coefficients)) {
  lm.cells <-
    tinydenseR::get.embedding(
      x = lm.cells,
      .full.model = "contrast.Source_TSO.binary",
      .contrast.of.interest = contrast,
      .verbose = TRUE
    )
}

# Visualize pePC1 for the early CRIT vs MILD contrast
(tinydenseR::plotSampleEmbedding(
  x = lm.cells,
  .embedding = "pePC",
  .color.by = "Source",
  .sup.embed.slot = "CRIT_earlyVSMILD_early",
  .x.by = "Source_TSO.binary"
) +
     ggplot2::theme(axis.text.x = ggplot2::element_text(vjust = 1,
                                                        hjust = 1,
                                                        angle = 30),
                    axis.title.x = ggplot2::element_blank()))

# Same plot, just highlighting the contrast groups
lm.cells$metadata |>
  cbind(lm.cells$sample.embed$pepc$CRIT_earlyVSMILD_early$coord) |>
  (\(x)
   ggplot2::ggplot(data = x,
                   mapping = ggplot2::aes(x = Source_TSO.binary,
                                          y = pePC1)) +
     ggplot2::theme_bw() +
     ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5),
                    axis.text.x = ggplot2::element_text(vjust = 1,
                                                        hjust = 1,
                                                        angle = 30),
                    axis.title.x = ggplot2::element_blank(),
                    legend.title = ggplot2::element_blank()) +
     ggplot2::guides(color = ggplot2::guide_legend(override.aes = list(size = 5))) +
     ggplot2::scale_color_manual(values = unname(obj = tinydenseR::Color.Palette[1,c(5,1:3)])) +
     ggplot2::geom_boxplot(outliers = FALSE,
                           color = 
                             (levels(x = x$Source_TSO.binary)  %in% 
                             c("CRIT_early",
                               "MILD_early")) |>
                             ifelse(yes = "black",
                                          no = "grey90")) +
     ggplot2::geom_point(mapping = ggplot2::aes(color = Source),
                         size = I(x = 1),
                         position = ggplot2::position_jitter(height = 0,
                                                             width = 0.2,
                                                             seed = 123),
                         alpha = ifelse(test = x$Source_TSO.binary %in% 
                                          c("CRIT_early",
                                            "MILD_early"),
                                        yes = 1,
                                        no = 0.25)) +
     ggplot2::labs(title = "early: CRIT vs MILD",
                   y = paste0("pePC1 (",
                              round(x = lm.cells$sample.embed$pepc$CRIT_earlyVSMILD_early$perc.tot.var.exp,
                                    digits = 1),
                              "%)")) +
     ggh4x::force_panelsizes(rows = grid::unit(x = 2,
                                               units = "in"),
                             col = grid::unit(x = 2,
                                              units = "in"))
  )()

```

### Beeswarm Summary

`plotBeeswarm()` provides a compact overview of which landmark clusters are
significantly enriched or depleted across multiple contrasts simultaneously.

```{r beeswarm, fig.width=10, fig.height=4}
tinydenseR::plotBeeswarm(
    x = lm.cells,
    .model.name = "contrast.Source_TSO.binary",
    .coefs = c("earlyInfectionVSuninfected",
               "lateInfectionVSearlyInfection",
               "CRIT_earlyVSMILD_early"),
    .label.substr.rm = "group",
    .split.by = "clustering",
    .perc.plot = FALSE,
    .q.from = "pca.weighted.q",
    .swarm.title = "COMBAT landmarks") |> 
  (\(x)
   x +
     ggplot2::facet_grid(cols = ggplot2::vars(facets), 
                         scales = "fixed", 
                         labeller = ggplot2::as_labeller(
                           c(
                             "earlyInfectionVSuninfected" = "early COVID vs healthy",
                             "lateInfectionVSearlyInfection" = "COVID: late vs early",
                             "CRIT_earlyVSMILD_early" = "early: CRIT vs MILD"
                           ))) +
     ggh4x::force_panelsizes(cols = ggplot2::unit(x = 1.5,
                                                  units = "in"),
                             rows = ggplot2::unit(x = (unique(x = x$data$split.by) |>
                                                         length()) * 0.2,
                                                  units = "in"))
  )()
```

## Step 4: Decomposing the Density Contrast with `get.plsDE()`

Differential density analysis tells us where changes occur across the
landmark landscape, but not what molecular features drive those changes.
`get.plsDE()` bridges this gap by decomposing the density contrast into PLS
components whose loadings reveal the genes underlying the signal.

We focus on the early CRIT vs MILD contrast, which captures the most
clinically actionable comparison — the cellular features that distinguish
patients who progress to critical illness from those with mild disease at an
early time point.

```{r run-plsDE}
lm.cells <-
  tinydenseR::get.plsDE(
    x = lm.cells,
    .coef.col = "CRIT_earlyVSMILD_early",
    .model.name = "contrast.Source_TSO.binary",
    .verbose = TRUE
  )
```

### plsDE1: Dysregulated Monocyte States

plsDE1 captures the dominant signal distinguishing early critical from mild
COVID-19. Examining its scores and loadings reveals that classical monocytes
are at the center of this dysregulation.

```{r plsDE1-scores, fig.width=10, fig.height=4}
tinydenseR::plotPlsDE(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsDE.dim = 1,
  .point.size = 0.1
)
```

The left panel shows landmark scores: landmarks in the monocyte region carry
the strongest positive plsDE1 scores, indicating that this component is driven
by monocyte-state variation. The right panel confirms tight coupling between
plsDE1 scores and the density contrast (Y).

The plsDE1 gene loadings identify the molecular signatures underlying
this signal. Positive loadings indicate features upregulated in the direction
of the contrast (CRIT > MILD), while negative loadings indicate
downregulation.

```{r plsDE1-heatmap, fig.width=8, fig.height=6}
tinydenseR::plotPlsDEHeatmap(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsDE.dim = 1,
  .order.by = "plsDE.dim",
  .feature.font.size = I(x = 5)
)
```

#### Interpretation

plsDE1 loadings highlight pro-inflammatory and innate immune genes
being upregulated in critical cases — consistent with the hyperinflammatory
monocyte phenotype that has been reported in critical COVID-19. Negative
loadings correspond to genes involved in antigen presentation and
classical monocyte identity (e.g., HLA class II genes), reflecting the
loss of antigen-presenting capacity previously documented in critical
patients.

### plsDE2: Expanded Immature Cells and Depleted pDCs

plsDE2 captures a secondary axis of variation. Inspecting the positive scores
reveals that this component is driven by landmarks in regions annotated as
progenitor or immature cell populations — consistent with emergency
hematopoiesis and the well-documented expansion of immature myeloid cells
in critical COVID-19 PBMC. The most negative scores correspond to landmarks in the plasmacytoid dendritic cell (pDC) region, which are known to be depleted in critical COVID-19.

```{r plsDE2-scores, fig.width=10, fig.height=4}
tinydenseR::plotPlsDE(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsDE.dim = 2,
  .point.size = 0.1
)
```

```{r plsDE2-heatmap, fig.width=8, fig.height=6}
tinydenseR::plotPlsDEHeatmap(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsDE.dim = 2,
  .order.by = "plsDE.dim",
  .feature.font.size = I(x = 5)
)
```

#### Interpretation

plsDE2 positive loadings highlight genes associated with immature or progenitor cell states (e.g., stem cell markers, cell-cycle genes) being positively loaded,
reflecting the expansion of these populations in critical COVID-19. Negative loading on the other hand are markers of plasmacytoid dendritic cells (pDCs) and, to a lesser extent, effector memory T cells, highlighting their depletion early in critical COVID-19. These findings aligns with reports of the role pDCs in most severe COVID-19 cases, and, more importantly, emergency myelopoiesis producing immature, dysfunctional cells that enter circulation during critical infection.

## Summary

This analysis demonstrates the full `tinydenseR` workflow on a real clinical
dataset:

| Step | Function | Purpose |
|------|----------|---------|
| 1 | `RunTDR()` | Build landmarks, graph, and density profiles |
| 2 | `plotUMAP()`, `plotSampleEmbedding()` | Validate biological structure |
| 3 | `get.lm()` with contrasts | Differential density testing, controlling for sex, age, and time since onset |
| 4 | `get.plsDE()` | Decompose density contrast into interpretable gene programs |

The key biological findings are:

- plsDE1 identifies dysregulated monocyte states in early critical
  COVID-19, with loadings pointing to pro-inflammatory activation and loss of
  antigen-presenting function in classical monocytes.
- plsDE2 reveals expanded immature cell populations, consistent with emergency hematopoiesis, and depleted pDCs in critical cases.

These results are obtained from a single unified framework that handles
differential density analysis while controlling for covariates through the
contrast design in `get.lm()`, followed by molecular decomposition via
`get.plsDE()`.

## Session Info

```{r session-info}
sessionInfo()
```
