---
title: "COVID-19 PBMC Analysis with tinydenseR: Identifying Dysregulated Cell States"
author: "Pedro Milanez-Almeida"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{COVID-19 PBMC Analysis with tinydenseR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  out.width = "100%",
  eval = TRUE
)
```

## Introduction

This vignette demonstrates how `tinydenseR` can be applied to a real-world
clinical scRNA-seq dataset to identify dysregulated cell states in COVID-19
patients. We use PBMC data from the COMBAT (COvid-19 Multi-omics Blood
ATlas) consortium to show how differential density analysis combined with
PLS-based decomposition (plsD) reveals:

1. **A monocyte-centered transcriptional axis** separating neutrophil
   degranulation-related inflammatory features from interferon alpha/gamma
   response programs in early critical vs. mild disease — captured by plsD
   component 1 (plsD1).
2. **A secondary severity-associated axis of hematopoietic stress** beyond
   the dominant monocyte program, with progenitor-like cell-associated genes
   at one extreme and pDC- and NK cell-associated genes at the other —
   captured by plsD component 2 (plsD2).

### Analysis Overview

The core workflow follows four steps:

1. `RunTDR()` — build landmarks from an on-disk count matrix and compute
   sample-level density profiles.
2. Visualize landmark and sample embeddings — inspect the UMAP of landmarks
   and the sample PCA to confirm biological structure.
3. `get.lm()` with contrasts — perform differential density analysis while
   controlling for covariates (sex, age, time since symptom onset).
4. `get.plsD()` — decompose the density contrast into interpretable
   components that jointly capture differential abundance and differential
   expression.

### Data Source

The COMBAT dataset is available from CELLxGENE
([link](https://datasets.cellxgene.cziscience.com/687c09ff-731a-4e3d-ac07-4c29c33a6338.h5ad)). The original study is described in:

> COMBAT Consortium. *A blood atlas of COVID-19 defines hallmarks of disease
> severity and specificity.* Cell 185, 916–938 (2022).

## Setup

### Load Packages

```{r load-packages}
library(tinydenseR)
library(dplyr)
library(forcats)
library(ggplot2)
library(patchwork)
library(limma)
library(anndataR)
library(BPCells)
```

### Load and Prepare Data

We read the COMBAT `.h5ad` file, keeping data on disk via `anndataR` and
`BPCells` for memory efficiency. This section shows the key preprocessing
steps; adjust file paths to match your local setup.

```{r load-data}
# Path to the downloaded h5ad file
file_path <- "~/sandbox/COVID19_COMBAT/derived_data/COMBAT.h5ad"

# Access h5ad keeping data on disk
h5ad <- 
  anndataR::read_h5ad(
    path = file_path,
    as = "HDF5AnnData")

# Extract metadata and feature information
feature_metadata <- h5ad$var
metadata <- h5ad$obs
```

### Quality Control and Filtering

We apply basic QC thresholds and restrict the analysis to healthy volunteers
(HV) and COVID-19 patients with mild, severe, or critical disease. Samples
with fewer than 200 cells are excluded.

```{r qc-filter}
# Basic QC: minimum 1000 UMI and 500 genes per cell
keep.QC.cells <- 
  (metadata$QC_total_UMI >= 1000) & 
  (metadata$QC_ngenes >= 500)

# Keep only HV, MILD, SEV, and CRIT patients with >= 200 cells
smpl.to.keep <-
  metadata$scRNASeq_sample_ID[
    metadata$Source %in% c("HV", "COVID_MILD", "COVID_SEV", "COVID_CRIT")
  ] |> 
  table() |>
  (\(x) names(x[x >= 200]))()

metadata <-
  metadata[keep.QC.cells & (metadata$scRNASeq_sample_ID %in% smpl.to.keep), ]
```

### Prepare the On-Disk Count Matrix

We use `BPCells` to store the count matrix on disk and subset it to QC-passing
cells with unique gene symbols.

```{r ondisk-matrix}
# Remove duplicated gene symbols
keep.unique.genes <- !duplicated(feature_metadata$feature_name)
feature_metadata <- feature_metadata[keep.unique.genes, ] |> droplevels()

# Write BPCells on-disk matrix
ondisk_path <- 
  "~/sandbox/COVID19_COMBAT/derived_data/COMBAT_COVID_ondisk/"

ondisk.data <-
  BPCells::open_matrix_anndata_hdf5(
    path = file_path,
    group = "/X")

if(!file.exists(path = ondisk_path)){
  # write this BPCells matrix to disk
  BPCells::write_matrix_dir(
    mat = ondisk.data, 
    dir = ondisk_path)
}

# Re-open and subset
ondisk.mat <-
  BPCells::open_matrix_dir(dir = ondisk_path)

ondisk.mat <-
  ondisk.mat[keep.unique.genes, keep.QC.cells & (metadata$scRNASeq_sample_ID %in% smpl.to.keep)]

rownames(x = ondisk.mat) <- feature_metadata$feature_name
```

### Prepare Metadata Covariates

We encode disease severity as ordered factor levels and binarize time since
symptom onset into "early" and "late" to allow for non-linear effects. We
also create a combined `Source_TSO.binary` variable for the contrast design.

```{r prepare-metadata}
metadata <-
  metadata |>
  dplyr::mutate(
    Source = factor(
      x = Source,
      levels = c("HV", "COVID_MILD", "COVID_SEV", "COVID_CRIT")),
    TimeSinceOnset = ifelse(
      test = Source != "HV", 
      yes = TimeSinceOnset, 
      no = -1)
  ) |>
  droplevels()
```

## Step 1: Build Landmarks with `RunTDR()`

`RunTDR()` is the main entry point. It creates landmark cells, builds a
neighborhood graph, computes sample-level density profiles, and maps cell type
annotations — all in a single call.

```{r run-tdr}
set.seed(seed = 123) # for reproducibility
lm.cells <-
  tinydenseR::RunTDR(
    x = ondisk.mat,
    .sample.var = "scRNASeq_sample_ID", 
    .cell.meta = metadata,
    .assay.type = "RNA",
    .celltype.vec = "cell_type",
    .label.confidence = 0.5,
    .verbose = FALSE
  )
```

## Step 2: Visualize Landmark and Sample Embeddings

Before statistical testing, we inspect the landmark UMAP and sample PCA to
confirm that the data exhibits the expected biological structure.

```{r umap-celltype, fig.width=10, fig.height=4}
tinydenseR::plotUMAP(
  x = lm.cells,
  .feature = lm.cells$landmark.annot$celltyping$ids,
  .plot.title = "COMBAT landmarks (cell types)"
)  +
  ggplot2::guides(color = ggplot2::guide_legend(
    override.aes = list(size = 5),
    ncol = 2))
```

We can also project individual gene expression (e.g., canonical markers) onto
the UMAP to validate cell identities:

```{r umap-markers, fig.width=10, fig.height=4}
# Classical monocyte marker
p1 <- 
  tinydenseR::plotUMAP(
  x = lm.cells,
  .feature = lm.cells$landmarks[, "LYZ"],
  .plot.title = "LYZ (classical monocytes)")
# MHC class II
p2 <- 
  tinydenseR::plotUMAP(
    x = lm.cells,
     .feature = lm.cells$landmarks[, "HLA-DRA"],
     .plot.title = "HLA-DRA (MHC class II)")
# T cell marker
p3 <- 
  tinydenseR::plotUMAP(
    x = lm.cells,
     .feature = lm.cells$landmarks[, "CD3E"],
     .plot.title = "CD3E (T cells)")

p1 | p2 | p3
```

### Sample PCA

The sample PCA provides a low-dimensional view of inter-sample variation based
on their density profiles across landmarks. The color encodes sample library
size (log10 total cells).

```{r sample-pca, fig.width=10, fig.height=4}
tinydenseR::plotSampleEmbedding(
  x = lm.cells,
  .embedding = "pca",
  .color.by = "log10.n.cells"
)
```

## Step 3: Differential Density Analysis with `get.lm()`

### Encoding Covariates

We binarize time since onset into "early" vs. "late" and create a composite
factor that crosses disease severity with timing. This allows us to define
biologically meaningful contrasts while controlling for confounders.

```{r encode-covariates}
# Binarize TimeSinceOnset
lm.cells@metadata <-
  dplyr::mutate(
    .data = lm.cells@metadata,
    TSO.binary = 
      dplyr::case_when(
        Source == "HV" ~ "uninfected",
        Source != "HV" ~ ifelse(
          TimeSinceOnset > median(TimeSinceOnset[Source != "HV"]),
          "late", "early")
      ) |>
      factor(levels = c("uninfected", "early", "late"))
  )

# Standardize Age encoding
lm.cells$metadata$Age <-
  make.names(lm.cells$metadata$Age) |>
  gsub(pattern = "^X", replacement = "Age.", fixed = FALSE)

# Create composite Source x TSO variable
lm.cells$metadata$Source_TSO.binary <-
  paste(lm.cells$metadata$Source,
        lm.cells$metadata$TSO.binary,
        sep = "_") |>
          gsub(pattern = "^HV_|^COVID_", 
          replacement = "",
           fixed = FALSE) |>
            factor(levels = c("uninfected", "MILD_early", "MILD_late", "SEV_early", "SEV_late", "CRIT_early", "CRIT_late"))
```

### Design and Contrast Matrices

We set up a cell-means design with `~ 0 + Source_TSO.binary + sex + Age`,
so that `sex` and `Age` are controlled for in every contrast. The contrast
matrix defines the biologically interesting comparisons.

```{r design-contrast}
# Design matrix (cell-means model controlling for sex and Age)
.design <- 
  model.matrix(~ 0 + Source_TSO.binary + sex + Age,
               data = lm.cells$metadata) |> 
  (\(x) {
    colnames(x) <- gsub("^Source_TSO.binary", "", colnames(x))
    x
  })()

# Contrast matrix
.contrast.matrix <- 
  limma::makeContrasts(
    # Critical vs mild within early timepoint
    CRIT_earlyVSMILD_early =
      CRIT_early - MILD_early,
    
    levels = .design
  )
```

### Fit the Model

`get.lm()` fits a limma model to the log-transformed fuzzy density profiles
across landmarks, applying the contrast matrix. This identifies which
landmarks are differentially abundant between conditions — analogous to
finding differentially expressed genes in bulk RNA-seq, but at the level of
cell density.

```{r fit-lm}
lm.cells <- 
  tinydenseR::get.lm(
    x = lm.cells,
    .design = .design,
    .contrasts = .contrast.matrix,
    .model.name = "contrast.Source_TSO.binary",
    .verbose = TRUE
  )
```

### Visualize Differential Density

We can now overlay the q-values and fold-change directions onto the landmark
UMAP. Density shifts span multiple immune compartments — including monocyte/DC,
T/NK, and B/plasmablast-rich regions of the manifold — indicating broad cell-state
remodeling in early critical relative to mild COVID-19.

```{r da-umap, fig.width=10, fig.height=4}
# get coloring based on fold-change direction and significance for each contrast
# Early CRIT vs MILD — the key contrast
tinydenseR::plotUMAP(
    x = lm.cells,
    .feature = lm.cells$results$lm$contrast.Source_TSO.binary$fit$coefficients[, "CRIT_earlyVSMILD_early"],
    .plot.title = "early: CRIT vs MILD",
    .cat.feature.color = unname(obj = Color.Palette[1, c(1, 6, 2)]),
    .color.label = "q < 0.1"
  ) +
  ggplot2::theme(plot.subtitle = ggplot2::element_blank())
```

### Sample Embeddings by Contrast

We compute a supervised sample embedding (partial-effect PC, pePC) for the
early CRIT vs MILD contrast. pePC1 provides a single quantitative score that
summarizes how each sample deviates along the contrast direction, with the
fraction of total density variance it explains reported on the y-axis.

```{r sample-embed, fig.width=10, fig.height=4}
# Compute sample embeddings for each contrast
lm.cells <-
    tinydenseR::get.embedding(
      x = lm.cells,
      .full.model = "contrast.Source_TSO.binary",
      .contrast.of.interest = "CRIT_earlyVSMILD_early",
      .verbose = TRUE
    )

# Visualize pePC1 for the early CRIT vs MILD contrast
tinydenseR::plotSampleEmbedding(
  x = lm.cells,
  .embedding = "pePC",
  .color.by = "Source",
  .sup.embed.slot = "CRIT_earlyVSMILD_early",
  .x.by = "Source_TSO.binary"
) +
     ggplot2::theme(axis.text.x = ggplot2::element_text(vjust = 1,
                                                        hjust = 1,
                                                        angle = 30),
                    axis.title.x = ggplot2::element_blank())

```

## Step 4: Decomposing the Density Contrast with `get.plsD()`

Differential density analysis tells us where changes occur across the
landmark landscape, but not what molecular features drive those changes.
`get.plsD()` bridges this gap by decomposing the density contrast into PLS
components whose loadings reveal the genes underlying the signal.

We focus on the early CRIT vs MILD contrast, which captures the most
clinically actionable comparison — the cellular features that distinguish
patients who progress to critical illness from those with mild disease at an
early time point.

```{r run-plsD}
lm.cells <-
  tinydenseR::get.plsD(
    x = lm.cells,
    .coef.col = "CRIT_earlyVSMILD_early",
    .model.name = "contrast.Source_TSO.binary",
    .verbose = TRUE,
    .min.prop = 0.001 # default is 0.005 but we lower it here to capture more subtle signals in this complex contrast
  )
```

### plsD1: Inflammatory Myeloid Program vs. Interferon Response

plsD1 captures the dominant signal distinguishing early critical from mild
COVID-19. Landmarks in the monocyte region carry the strongest plsD1 scores,
indicating that this component is driven by monocyte-state variation.

```{r plsD1-scores, fig.width=10, fig.height=4}
tinydenseR::plotPlsD(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsD.dim = 1
)
```

The left panel shows landmark scores: landmarks in the monocyte region carry
the strongest positive plsD1 scores, indicating that this component is driven
by monocyte-state variation. The right panel confirms tight coupling between
plsD1 scores and the density contrast (Y).

The plsD1 gene loadings identify the molecular signatures underlying
this signal. Positive loadings indicate features upregulated in the direction
of the contrast (CRIT > MILD), while negative loadings indicate
downregulation.

```{r plsD1-heatmap, fig.width=8, fig.height=6}
tinydenseR::plotPlsDHeatmap(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsD.dim = 1,
  .order.by = "plsD.dim",
  .feature.font.size = I(x = 5)
)
```

```{r plsD1-gsea, fig.width=10, fig.height=4}
# enrichment analysis setup
msigdb.H <-
  msigdbr::msigdbr(species = "Homo sapiens")

hypotheses.1 <-
  c(
    "REACTOME_NEUTROPHIL_DEGRANULATION",
    "HALLMARK_INTERFERON_ALPHA_RESPONSE",
    "HALLMARK_INTERFERON_GAMMA_RESPONSE"  
  ) |> 
  (\(x)
   unique(x = x[x %in% msigdb.H$gs_name])
   )()
  
msigdb.list.1 <-
  msigdb.H[msigdb.H$gs_name %in% hypotheses.1,
           c("gene_symbol",
             "gs_name")] |>
  (\(x)
   split(x = x$gene_symbol,
         f = x$gs_name)
  )()  |>
  lapply(FUN = unique)

set.seed(seed = 123)
fgseaRes.plsD1 <- 
  fgsea::fgseaMultilevel(pathways = msigdb.list.1, 
                         stats = lm.cells$results$pls$CRIT_earlyVSMILD_early$loadings[,"plsD1"],
                         eps = 1e-50,
                         minSize = 5,
                         maxSize = 500,
                         nPermSimple = 1e5)

# add concordance score for leading edge genes (the higher the score, the higher the confidence in the results)
fgseaRes.plsD1$concordance <-
  lapply(X = fgseaRes.plsD1$leadingEdge,
         FUN = function(lead.edge){
           mean(x = lm.cells$results$pls$CRIT_earlyVSMILD_early$concordance.weights[lead.edge,"plsD1"])
         }) |>
  unlist()

fgseaRes.plsD1[order(-log10(x = padj)*sign(NES)),]

(p <-
  names(x = msigdb.list.1) |>
  lapply(FUN = function(pw) {
    
    fgsea::plotEnrichment(pathway = msigdb.list.1[[pw]],
                          stats =  lm.cells$results$pls$CRIT_earlyVSMILD_early$loadings[,"plsD1"]) +
      ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5,
                                                        size = I(x = 12)),
                     plot.subtitle = ggplot2::element_text(hjust = 0.5,
                                                           size = I(x = 10)),
      ) +
      ggplot2::labs(
        title = gsub(pattern = "_",
                                 replacement = " ",
                                 x = pw,
                                 fixed = TRUE) |>
            gsub(pattern = "^HALLMARK |^REACTOME ",
                    replacement = "",
                  fixed = FALSE) |>
                      tolower(),
                    subtitle = paste0("p = ",
                                      formatC(x = fgseaRes.plsD1[pathway == pw, padj],
                                              digits = 0,
                                              format = "e"))) +
      ggh4x::force_panelsizes(rows = grid::unit(x = 1,
                                                units = "in"),
                              col = grid::unit(x = 1.5,
                                               units = "in"))
  }) |>
  c(nrow = 1) |>
    do.call(what = gridExtra::grid.arrange))
```
#### Interpretation

Positive plsD1 loadings are enriched for **neutrophil degranulation**, with
key contributing genes including S100A12, CLU, VCAN, and CTSD. Negative
loadings are enriched for **interferon gamma response** and **interferon
alpha response** programs, driven by genes such as IFIT3, IFIT1, IFITM1,
HLA-DPA1, and HLA-DPB1. This axis is consistent with a shift from
interferon-responsive classical monocytes in early mild disease toward a
more inflammatory, dysregulated myeloid program in early critical disease,
in line with previous reports of myeloid dysregulation in severe COVID-19.

### plsD2: Progenitor-like vs. pDC/NK Programs

plsD2 captures a secondary axis of variation within the same contrast.
plsD2 scores are most extreme within two very small subsets of landmarks,
with positive scores mapping to progenitor-like regions and the most negative
scores mapping to plasmacytoid dendritic cell (pDC) and NK cell regions.

```{r plsD2-scores, fig.width=10, fig.height=4}
tinydenseR::plotPlsD(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsD.dim = 2
)
```

```{r plsD2-heatmap, fig.width=8, fig.height=6}
tinydenseR::plotPlsDHeatmap(
  x = lm.cells,
  .coef.col = "CRIT_earlyVSMILD_early",
  .plsD.dim = 2,
  .order.by = "plsD.dim",
  .feature.font.size = I(x = 5)
)
```

```{r plsD2-gsea, fig.width=10, fig.height=4}
hypotheses.2 <-
     c(
      "HE_LIM_SUN_FETAL_LUNG_C2_HSC_ELP_CELL",
      "KEGG_NATURAL_KILLER_CELL_MEDIATED_CYTOTOXICITY",
      "HE_LIM_SUN_FETAL_LUNG_C2_PDC_CELL"
      ) |> 
  (\(x)
   unique(x = x[x %in% msigdb.H$gs_name])
   )()
  
msigdb.list.2 <-
  msigdb.H[msigdb.H$gs_name %in% hypotheses.2,
           c("gene_symbol",
             "gs_name")] |>
  (\(x)
   split(x = x$gene_symbol,
         f = x$gs_name)
  )()  |>
  lapply(FUN = unique)

set.seed(seed = 123)
fgseaRes.plsD2 <- 
  fgsea::fgseaMultilevel(pathways = msigdb.list.2, 
                         stats = lm.cells$results$pls$CRIT_earlyVSMILD_early$loadings[,"plsD2"],
                         eps = 1e-50,
                         minSize = 5,
                         maxSize = 500,
                         nPermSimple = 1e5)

# add concordance score for leading edge genes (the higher the score, the higher the confidence in the results)
fgseaRes.plsD2$concordance <-
  lapply(X = fgseaRes.plsD2$leadingEdge,
         FUN = function(lead.edge){
           mean(x = lm.cells$results$pls$CRIT_earlyVSMILD_early$concordance.weights[lead.edge,"plsD2"])
         }) |>
  unlist()

fgseaRes.plsD2[order(-log10(x = padj)*sign(NES)),]

(p <-
  names(x = msigdb.list.2) |>
  lapply(FUN = function(pw) {
    
    fgsea::plotEnrichment(pathway = msigdb.list.2[[pw]],
                          stats =  lm.cells$results$pls$CRIT_earlyVSMILD_early$loadings[,"plsD2"]) +
      ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5,
                                                        size = I(x = 12)),
                     plot.subtitle = ggplot2::element_text(hjust = 0.5,
                                                           size = I(x = 10)),
      ) +
      ggplot2::labs(
        title = gsub(pattern = "_",
                                 replacement = " ",
                                 x = pw,
                                 fixed = TRUE) |>
            gsub(pattern = "^KEGG |^HE LIM SUN FETAL LUNG C2 ",
                    replacement = "",
                  fixed = FALSE) |>
                      tolower(),
                    subtitle = paste0("p = ",
                                      formatC(x = fgseaRes.plsD2[pathway == pw, padj],
                                              digits = 0,
                                              format = "e"))) +
      ggh4x::force_panelsizes(rows = grid::unit(x = 1,
                                                units = "in"),
                              col = grid::unit(x = 1.5,
                                               units = "in"))
  }) |>
  c(nrow = 1) |>
    do.call(what = gridExtra::grid.arrange))

rank(x = lm.cells$results$pls$CRIT_earlyVSMILD_early$loadings[,"plsD2"])[c("NKG7", "PRF1", "CST7", "GZMB", "CCL5", "GZMH", "GZMA")]

```

#### Interpretation

Analysis of plsD2 loadings reveals the progenitor-like cell-associated genes
CD34, CYTL1 and PRSS57 to be highly expressed in landmarks with the highest
plsD2 scores, while the pDC-associated genes CLEC4C, LILRA4 and TPM2 are
highly expressed in landmarks with low plsD2 scores. Furthermore, the NK
cell-associated genes NKG7, PRF1, CST7, GZMB, CCL5, GZMH and GZMA are among
the 100 genes with the most negative loadings. These findings are consistent
with reports that severe COVID-19 is accompanied by emergency myelopoiesis
and impaired antiviral innate immunity, including expansion of proliferative
progenitor/myeloid states and reduced or dysfunctional pDC and NK cell
programs, suggesting that plsD2 captures a secondary severity-associated axis
of hematopoietic stress beyond the dominant dysregulated monocyte program
highlighted by plsD1.

## Summary

This analysis demonstrates the full `tinydenseR` workflow on a real clinical
dataset:

| Step | Function | Purpose |
|------|----------|---------|
| 1 | `RunTDR()` | Build landmarks, graph, and density profiles |
| 2 | `plotUMAP()`, `plotSampleEmbedding()` | Validate biological structure |
| 3 | `get.lm()` with contrasts | Differential density testing, controlling for sex, age, and time since onset |
| 4 | `get.plsD()` | Decompose density contrast into interpretable gene programs |

The key biological findings are:

- **plsD1** identifies a monocyte-centered transcriptional axis separating
  neutrophil degranulation-related inflammatory features from interferon
  alpha/gamma response programs in early critical vs. mild COVID-19.
- **plsD2** resolves a secondary severity-associated axis of hematopoietic
  stress, with progenitor-like cell-associated genes (CD34, CYTL1, PRSS57)
  at one extreme and pDC- (CLEC4C, LILRA4, TPM2) and NK cell-associated
  programs at the other, consistent with emergency myelopoiesis and impaired
  antiviral innate immunity in critical cases.

Together, these results illustrate how `tinydenseR` combines sample-level
modeling with exploratory decomposition to resolve multiple layers of
biological heterogeneity within a single clinically relevant contrast.

## Session Info

```{r session-info}
sessionInfo()
```
