---
title: "Differential Expression Analysis with tinydenseR"
author: "Pedro Milanez-Almeida"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Differential Expression Analysis with tinydenseR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  out.width = "100%",
  eval = TRUE
)
```

## Introduction

This vignette demonstrates how `tinydenseR` detects differential expression
in a simulated flow cytometry experiment. A target population (5% of cells,
held constant across groups) undergoes a very small activation-dependent
upregulation of Marker2 (log-mean shift from 0 to 0.5) in the treatment
group. Because target-cell abundance is fixed, systematic treatment
differences arise from within-population expression changes rather than
compositional shifts. Five markers are measured: three (Marker1, Marker4,
Marker5) define the target population, Marker2 carries the differential
expression signal, and Marker3 is non-informative. Marker-specific batch
effects are present.

The workflow follows four steps:

1. `RunTDR()` — build landmarks, graph, and density profiles from FCS files.
2. `get.lm()` — differential density testing to identify affected
   subpopulations, controlling for batch.
3. `get.plsD()` — decompose the density contrast; positive plsD1 loadings
   for Marker2 reflect the ground-truth activation.
4. `get.pbDE()` (design-mode) — pseudobulk DE within the affected cluster to
   confirm Marker2 as the most differentially expressed feature.

## Setup

```{r load-packages}
library(tinydenseR)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(flowCore)
library(flowWorkspace)

set.seed(seed = 42)
```

## Step 1: Simulate Data

We use `simulate_DE_data()` to generate FCS files with a single effect size
(Marker2 log-mean 0 → 0.5 in target cells upon activation), six samples per
group, and approximately 50 000 cells per sample.

```{r simulate-data}
sim_data <-
  tinydenseR::simulate_DE_data(
    sd_shifts = 0.5,
    seed = 42
  )

# Sample and cell metadata
sample_meta_DE <- sim_data$sample_meta
cell_meta_DE <- sim_data$cell_meta
```

### Verify Ground Truth Expression Change

We verify the simulated expression shift by matching FCS expression data
(already log-transformed by `simulate_DE_data()`) with ground truth cell
type labels from `cell_meta`:

```{r ground-truth-expression}
# Build per-cell ground truth by matching FCS rows with cell_meta row-by-row
ground_truth_DE <-
  lapply(X = seq_len(nrow(sample_meta_DE)),
         FUN = function(i) {
           sample_id <- sample_meta_DE$Sample[i]
           frame <- flowCore::read.FCS(sample_meta_DE$fcs_path[i])
           expr <- flowCore::exprs(frame)  # already log-transformed
           
           # cell_meta rows for this sample are in the same order as FCS rows
           cell_rows <- cell_meta_DE[cell_meta_DE$Sample == sample_id, ]
           
           data.frame(
             Sample = sample_id,
             Treatment = sample_meta_DE$Treatment[i],
             Batch = sample_meta_DE$Batch[i],
             CellType = cell_rows$CellType,
             Marker2 = expr[, "Marker2"]
           )
         }) |>
  dplyr::bind_rows()

# Linear model by cell type
DE_res_DE <-
  ground_truth_DE |>
  dplyr::group_by(Sample, Treatment, Batch, CellType) |>
  dplyr::summarize(Marker2 = mean(Marker2), .groups = "drop") |>
  dplyr::mutate(Treatment = factor(Treatment,
                                   levels = c("Baseline", "Activation"))) |>
  split(f = ~CellType) |>
  lapply(FUN = function(x) {
    mod <- lm(Marker2 ~ Treatment + Batch, data = x)
    data.frame(
      Marker = "Marker2",
      coef = unname(coef(mod)["TreatmentActivation"]),
      CI_lower = confint(mod)["TreatmentActivation", "2.5 %"],
      CI_upper = confint(mod)["TreatmentActivation", "97.5 %"]
    )
  }) |>
  dplyr::bind_rows(.id = "CellType")

ggplot2::ggplot(data = DE_res_DE,
                mapping = ggplot2::aes(x = coef, y = CellType)) +
  ggplot2::theme_bw() +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dotted", color = "blue") +
  ggplot2::geom_vline(xintercept = 0.5, linetype = "dashed", color = "red",
                      alpha = 0.5) +
  ggplot2::geom_errorbar(mapping = ggplot2::aes(xmin = CI_lower,
                                                xmax = CI_upper),
                         width = 0.2) +
  ggplot2::labs(title = "Ground Truth Differential Expression",
                x = "log Fold Change in Expression (Marker2)",
                y = "") +
  ggh4x::force_panelsizes(cols = grid::unit(x = 3, units = "in"),
                          rows = grid::unit(x = 2, units = "in"))
```

The shift is confined to target cells (red dashed line at 0.5); the "other"
population shows no change.

## Step 2: Build Landmarks with `RunTDR()`

We load the FCS files into a `cytoset`, attach sample-level metadata, and
call `RunTDR()`.

```{r prepare-cytoset}
# Sample metadata for the 0.5SD setting
.setting.meta <-
  sim_data$sample_meta |>
  dplyr::filter(SD_Shift == "0.5SD")

# Load FCS files into a cytoset
cs.DE <-
  flowWorkspace::load_cytoset_from_fcs(
    files = stats::setNames(.setting.meta$fcs_path,
                            .setting.meta$Sample))

# Attach Treatment and Batch to phenoData
flowWorkspace::pData(cs.DE)$Sample <-
  flowWorkspace::sampleNames(cs.DE)

flowWorkspace::pData(cs.DE)$Treatment <-
  .setting.meta$Treatment[
    match(flowWorkspace::sampleNames(cs.DE),
          paste0(.setting.meta$Sample, ".fcs"))]

flowWorkspace::pData(cs.DE)$Batch <-
  .setting.meta$Batch[
    match(flowWorkspace::sampleNames(cs.DE),
          paste0(.setting.meta$Sample, ".fcs"))]
```

```{r run-tdr}
set.seed(seed = 123)
lm.cells <-
  tinydenseR::RunTDR(
    x = cs.DE,
    .sample.var = "Sample",
    .assay.type = "cyto",
    .markers = paste0("Marker", 1:5),
    .seed = 123,
    .verbose = TRUE,
    .cl.resolution.parameter = 0.5
  )
```

### Sample PCA

The unsupervised sample PCA shows inter-sample variation based on density
profiles across landmarks, colored by treatment group:

```{r sample-pca, fig.width=5, fig.height=3}
smpl.pca <-
  tinydenseR::plotSampleEmbedding(
    x = lm.cells,
    .embedding = "pca",
    .color.by = "Treatment",
    .cat.feature.color = tinydenseR::Color.Palette[1, c(2, 1)],
    .panel.size = 1.5,
    .point.size = 3
  ) +
  ggplot2::labs(title = "PCA") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5),
                 legend.position = "bottom")

smpl.pca
```

## Step 3: Differential Density Analysis with `get.lm()`

We set up a design matrix controlling for batch. Even though the target
population abundance is constant, the within-population expression shift
creates a density redistribution across landmarks that `get.lm()` can detect.

```{r design-matrix}
# Ensure Treatment factor levels are ordered correctly
lm.cells@metadata$Treatment <-
  factor(lm.cells@metadata$Treatment,
         levels = c("Baseline", "Activation"))

.design <-
  model.matrix(~ Treatment + Batch,
               data = lm.cells@metadata) |>
  (\(x) {
    colnames(x) <- gsub("^Treatment|^Batch", "", colnames(x))
    x
  })()
```

```{r fit-lm}
lm.cells <-
  tinydenseR::get.lm(
    x = lm.cells,
    .design = .design
  )
```

### Supervised Sample Embedding

We fit reduced models and compute partial-effect PCs to produce a
quantitative per-sample score along the treatment effect.

```{r supervised-embedding}
# Reduced models
noTreatment.design <-
  model.matrix(~ Batch, data = lm.cells@metadata)

noBatch.design <-
  model.matrix(~ Treatment, data = lm.cells@metadata)

lm.cells <-
  tinydenseR::get.lm(
    x = lm.cells,
    .design = noTreatment.design,
    .model.name = "noTreatment",
    .verbose = TRUE
  ) |>
  tinydenseR::get.lm(
    .design = noBatch.design,
    .model.name = "noBatch",
    .verbose = TRUE
  )

# Compute partial-effect PCs
lm.cells <-
  tinydenseR::get.embedding(
    x = lm.cells,
    .full.model = "default",
    .red.model = "noTreatment",
    .term.of.interest = "Treatment",
    .verbose = FALSE
  ) |>
  tinydenseR::get.embedding(
    .full.model = "default",
    .red.model = "noBatch",
    .term.of.interest = "Batch",
    .verbose = FALSE
  )
```

```{r pepc-plot, fig.width=5, fig.height=3}
smpl.pePC <-
  tinydenseR::plotSampleEmbedding(
    x = lm.cells,
    .embedding = "pePC",
    .sup.embed.slot = "Treatment",
    .color.by = "Treatment",
    .cat.feature.color = tinydenseR::Color.Palette[1, c(2, 1)],
    .panel.size = 1.5,
    .point.size = 3
  ) +
  ggplot2::labs(title = "pePC") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5),
                 legend.position = "bottom")

(smpl.pca | smpl.pePC) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "bottom",
                 legend.justification = "center")
```

The pePC separates Baseline from Activation samples along a single supervised
axis, confirming that the expression-driven density redistribution is
captured after controlling for batch.

### Density Contrast and plsD

We visualize the density fold-change alongside the leading plsD component on
the landmark PCA.

```{r plsD, fig.width=8, fig.height=4}
# Density fold-change on landmark PCA
DE.dens <-
  tinydenseR::plotPCA(
    x = lm.cells,
    .feature = lm.cells@results$lm$default$fit$coefficients[, "Activation"],
    .plot.title = "Activation vs Baseline",
    .color.label = "density\nlog2(+0.5)FC",
    .panel.size = 1.5,
    .point.size = 1,
    .midpoint = 0
  ) +
  ggplot2::theme(plot.subtitle = ggplot2::element_blank())

# Run plsD decomposition
lm.cells <-
  tinydenseR::get.plsD(
    x = lm.cells,
    .coef.col = "Activation",
    .model.name = "default",
    .verbose = TRUE
  )

# plsD1 scores on landmark PCA
DE.plsD1 <-
  tinydenseR::plotPlsD(
    x = lm.cells,
    .coef.col = "Activation",
    .plsD.dim = 1,
    .embed = "pca",
    .panel.size = 1.5
  )[[1]]

# Patchwork: density contrast | plsD1
((DE.dens +
    ggplot2::guides(color = ggplot2::guide_colorbar(title.position = "top",
                                                    title.hjust = 0.5)) +
    ggplot2::labs(title = "") +
    ggplot2::theme(legend.position = "bottom",
                   legend.margin = ggplot2::margin(t = -0.1,
                                                   unit = "in"))) |
   (DE.plsD1 +
      ggplot2::labs(title = "") +
      ggplot2::theme(legend.margin = ggplot2::margin(t = 0.1,
                                                     unit = "in")))) +
  patchwork::plot_annotation(title = "Density contrast: Activation") &
  ggplot2::theme(
    plot.title =
      ggplot2::element_text(hjust = 0.5,
                            margin = ggplot2::margin(t = -0.1,
                                                     unit = "in")))
```

The left panel shows density log2 fold-change at each landmark: the
activation-induced expression shift in Marker2 redistributes target cells
across neighboring landmarks, producing a localized density contrast. The
right panel shows plsD1 scores, which co-localize with the density contrast.

### plsD Heatmap

The plsD heatmap shows landmark expression profiles ordered by plsD1 scores,
with marker loadings on the right. Positive loadings for Marker2 reflect the
ground-truth activation signal.

```{r plsD-heatmap, fig.width=8, fig.height=5}
tinydenseR::plotPlsDHeatmap(
  x = lm.cells,
  .coef.col = "Activation",
  .plsD.dim = 1,
  .order.by = "plsD.dim",
  .panel.height = 2,
  .feature.font.size = I(x = 8)
)
```

## Step 4: Pseudobulk DE with `get.pbDE()`

Design-mode pseudobulk DE compares treatment vs. baseline within the most
affected cluster to identify which markers are differentially expressed.

```{r clustering, fig.width=6, fig.height=4}
cl.plot <-
  tinydenseR::plotPCA(
    x = lm.cells,
    .feature = lm.cells@landmark.annot$clustering$ids,
    .plot.title = "clustering",
    .point.size = 1,
    .panel.size = 1.5,
    .legend.position = "bottom"
  ) |>
  (\(x)
   x +
     ggplot2::guides(color = ggplot2::guide_legend(
       ncol = 2,
       override.aes = list(size = I(x = 5)))) +
     ggplot2::theme(plot.subtitle = ggplot2::element_blank()) +
     ggrepel::geom_text_repel(
       data = x$data |>
         dplyr::group_by(feature) |>
         dplyr::summarize(PC1 = mean(PC1),
                          PC2 = mean(PC2),
                          .groups = "drop"),
       mapping = ggplot2::aes(label = feature),
       size = I(x = 4),
       color = "black",
       seed = 123,
       min.segment.length = -Inf)
  )()

cl.plot
```

```{r pbDE}
lm.cells <-
  tinydenseR::get.pbDE(
    x = lm.cells,
    .design = .design,
    .id = "cluster.4"
  )
```

```{r pbDE-plot, fig.width=7, fig.height=4}
pbDE.plot <-
  tinydenseR::plotPbDE(
    x = lm.cells,
    .population.name = "cluster.4",
    .coefs = "Activation",
    .row.space.scaler = 0.065,
    .col.space.scaler = 0.2,
    .order.by = "none"
  ) +
  ggplot2::labs(title = "cluster.4: Activation",
                subtitle = "pseudobulk") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(hjust = 1,
                                                     vjust = 1,
                                                     angle = 30),
                 axis.title = ggplot2::element_blank(),
                 axis.text.y = ggplot2::element_blank()) +
  ggplot2::guides(fill = ggplot2::guide_colorbar(position = "bottom",
                                                 title.position = "top",
                                                 title.hjust = 0.5)) +
  ggplot2::coord_flip()

cl.plot | pbDE.plot
```

Design-mode pseudobulk DE correctly identifies Marker2 as the most
differentially expressed feature within the affected cluster, confirming
recovery of the ground-truth activation signal.

## Summary

This analysis demonstrates the `tinydenseR` workflow on a simulated flow
cytometry differential expression experiment:

| Step | Function | Purpose |
|------|----------|---------|
| 1 | `RunTDR()` | Build landmarks, graph, and density profiles from FCS files |
| 2 | `get.lm()` | Differential density testing to identify affected subpopulations |
| 3 | `get.plsD()` | Decompose density contrast into interpretable marker programs |
| 4 | `get.pbDE()` | Design-mode pseudobulk DE within the affected cluster |

tinydenseR detected the very small activation-dependent upregulation of
Marker2 (log-mean 0 → 0.5) in a 5% target population, in the presence of
batch effects. Positive plsD1 loadings for Marker2 reflected the ground-truth
target activation, and design-mode pseudobulk DE within the most affected
cluster correctly identified Marker2 as the most differentially expressed
feature.

## Session Information

```{r session-info}
sessionInfo()
```
