---
title: "Differential Abundance Analysis with tinydenseR"
author: "Pedro Milanez-Almeida"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Differential Abundance Analysis with tinydenseR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  out.width = "100%"
)
```

## Introduction

Differential abundance (DA) analysis identifies cell populations whose proportions change between experimental conditions.

### What is Differential Abundance Analysis?

In single-cell experiments, we often want to know:

- Which cell populations are depleted or enriched in treatment vs. control?
- How do cell type proportions change across conditions?
- Are these changes statistically significant after accounting for technical variation?

Traditional approaches cluster cells first, then test for proportion differences. This can be problematic because:

1. **Clustering is subjective** - resolution parameters strongly affect results
2. **Biological variation is continuous** - hard boundaries between clusters may not reflect biology

### The tinydenseR Approach

`tinydenseR` uses a landmark-based framework that:

- Represents the cell population landscape with data-driven selection of landmark cells
- Maps all cells to their nearest landmarks
- Models abundance at each landmark using **samples as biological replicates**
- Detects regions of differential abundance without arbitrary clustering

This approach is particularly powerful for:

- scRNA-seq, flow, mass and spectral cytometry data
- Detecting subtle population shifts
- Accounting for batch effects and confounders
- Maintaining statistical rigor at scale

## Simulating Data with Differential Abundance

For this tutorial, we'll simulate flow cytometry data where a target cell population is depleted 2-fold in the "Depletion" condition compared to "Baseline". We'll simulate three scenarios with different baseline proportions (0.5%, 5%, and 50%) to verify our simulation design, but **focus our analysis on the most challenging case: the rare 0.5% population**. This demonstrates tinydenseR's ability to detect subtle changes in rare cell populations.

### Load Required Packages

```{r load-packages}
library(tinydenseR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(ggh4x)
library(Matrix)

set.seed(seed = 42)
```

### Simulation Parameters

```{r sim-params}
# Experimental design
groups <- c("Baseline", "Depletion")
batches <- c("Batch1", "Batch2")
settings <- c(0.005, 0.05, 0.5)  # Target cell proportions for Baseline
samples_per_group <- 6
mean_cells <- 50000
sd_cells <- 500  # Noise in total cell count
```

### Generate Simulated Data

We'll simulate 5 markers where:

- **Marker1, Marker4, Marker5**: Define the target cell population (high expression)
- **Marker2, Marker3**: Non-informative markers (same distribution across cell types)
- **Batch effects**: Multiplicative noise added to Batch2 samples

```{r simulate-data}
# Initialize storage
data_list_DA <- list()

# Simulate data
for (setting in settings) {
  for (group in groups) {
    for (sample_id in 1:samples_per_group) {
      batch <- if(sample_id %% 2 == 0) "Batch2" else "Batch1"
      
      total_cells <- base::round(x = stats::rnorm(n = 1, mean = mean_cells, sd = sd_cells))
      total_cells <- base::max(total_cells, 1000)  # Ensure minimum cells
      
      # Two-fold difference in proportions
      proportion <- if (group == "Baseline") setting else setting / 2
      
      num_interest <- base::round(x = total_cells * proportion)
      num_other <- total_cells - num_interest
      cell_types <- base::sample(x = c(base::rep(x = "target", times = num_interest), 
                                        base::rep(x = "other", times = num_other)))
      
      # Simulate expression data
      marker1 <- base::numeric(length = total_cells)
      marker2 <- stats::rlnorm(n = total_cells, meanlog = 0, sdlog = 1.5)
      marker3 <- stats::rlnorm(n = total_cells, meanlog = 0, sdlog = 2.5)
      marker4 <- base::numeric(length = total_cells)  
      marker5 <- base::numeric(length = total_cells)  
      
      # Assign marker expression based on cell type
      marker1[cell_types == "other"] <- stats::rlnorm(n = base::sum(cell_types == "other"), 
                                                       meanlog = 0, sdlog = 2)
      marker1[cell_types == "target"] <- stats::rlnorm(n = base::sum(cell_types == "target"), 
                                                        meanlog = 5, sdlog = 2)
      
      marker4[cell_types == "other"] <- stats::rlnorm(n = base::sum(cell_types == "other"), 
                                                       meanlog = 0, sdlog = 1.2)
      marker4[cell_types == "target"] <- stats::rlnorm(n = base::sum(cell_types == "target"), 
                                                        meanlog = 3, sdlog = 1.2)
      
      marker5[cell_types == "other"] <- stats::rlnorm(n = base::sum(cell_types == "other"), 
                                                       meanlog = 0, sdlog = 1.8)
      marker5[cell_types == "target"] <- stats::rlnorm(n = base::sum(cell_types == "target"), 
                                                        meanlog = 7, sdlog = 1.8)
      
      # Add batch effect
      if (batch == "Batch2") {
        marker1 <- marker1 * stats::rlnorm(n = total_cells, meanlog = 0.1, sdlog = 0.5)
        marker2 <- marker2 * stats::rlnorm(n = total_cells, meanlog = 0.2, sdlog = 0.3)
        marker3 <- marker3 * stats::rlnorm(n = total_cells, meanlog = 0.3, sdlog = 0.4)
        marker4 <- marker4 * stats::rlnorm(n = total_cells, meanlog = 0.4, sdlog = 0.3)
        marker5 <- marker5 * stats::rlnorm(n = total_cells, meanlog = 0.5, sdlog = 0.5)
      }
      
      sample_name <- base::paste0(group, "_S", sample_id, "_Set", setting * 100)
      df <- base::data.frame(
        Sample = sample_name,
        Treatment = group,
        Batch = batch,
        Setting = base::paste0(setting * 100, "%"),
        CellType = cell_types,
        Marker1 = marker1,
        Marker2 = marker2,
        Marker3 = marker3,
        Marker4 = marker4,
        Marker5 = marker5
      )
      data_list_DA[[base::length(x = data_list_DA) + 1]] <- df
    }
  }
}

# Combine all samples
final_data_DA <- base::do.call(what = rbind, args = data_list_DA)
```

### Visualize Data Structure

Let's take a look at the data structure. Please note marker expression profiles by cell type showing target population is defined by Marker1, Marker4, and Marker5.

```{r viz-markers}
final_data_DA |>
    dplyr::filter(Setting == "0.5%") |>
    tidyr::pivot_longer(
        cols = c(Marker1, Marker2, Marker3, Marker4, Marker5),
        names_to = "Marker",
        values_to = "value"
    ) |>
    ggplot2::ggplot(mapping = ggplot2::aes(x = Marker, 
                                           y = value)) +
    ggplot2::facet_grid(cols = ggplot2::vars(CellType)) +
    ggplot2::theme_bw() +
    ggplot2::theme(
        plot.title = ggplot2::element_text(hjust = 0.5),
        legend.position = "right",
        axis.text.x = ggplot2::element_text(hjust = 1, vjust = 1, angle = 30)
    ) +
    ggplot2::labs(
        title = "Marker Expression by Cell Type",
        x = "", 
        y = "Expression Level"
    ) +
  ggplot2::geom_hline(yintercept = 1,
                      linetype = "dashed",
                      color = "red") +
    ggplot2::scale_y_continuous(transform = "log10",
                                labels = scales::trans_format(trans = "log10", 
                                                              format = scales::math_format(10^.x))) + 
    ggplot2::annotation_logticks(sides = "l") +
    ggplot2::geom_violin(color = "black",
                         alpha = 0,
                         draw_quantiles = 0.5) +
    ggh4x::force_panelsizes(cols = grid::unit(x = 2, units = "in"), 
                            rows = grid::unit(x = 2, units = "in"))
```

```{r viz-ground-truth}
# pseudo-colors
pseudo.colors <-
    grDevices::colorRampPalette(
        colors = c(
            "blue",
            "green",
            "yellow",
            "red"
        ))

# Example: generate 256 colors
pseudo.palette <-
  pseudo.colors(n = 256)

# Example
final_data_DA |>
    dplyr::filter(Sample %in% c("Baseline_S1_Set0.5", "Depletion_S1_Set0.5")) |>
    ggplot2::ggplot(mapping = ggplot2::aes(x = Marker1,
                                           y = Marker5)) +
    ggplot2::facet_grid(cols = ggplot2::vars(Sample)) +
    ggplot2::theme_bw() +
    ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5), 
                   legend.position = "none") +
    ggplot2::labs(title = "Example Marker Expression in Two Different Samples") +
    ggplot2::scale_x_continuous(transform = "log10",
                                labels = scales::trans_format(trans = "log10", 
                                                              format = scales::math_format(10^.x))) + 
    ggplot2::scale_y_continuous(transform = "log10",
                                labels = scales::trans_format(trans = "log10", 
                                                              format = scales::math_format(10^.x))) + 
    ggplot2::annotation_logticks(sides = "lb") +
    ggplot2::scale_fill_gradientn(colors = pseudo.palette,
                                  transform = "log10") +
    ggplot2::stat_bin_hex(bins = 64) + 
    ggh4x::force_panelsizes(cols = grid::unit(x = 2, units = "in"), 
                            rows = grid::unit(x = 2, units = "in"))
```

Below, target cells (shown in purple) form a distinct subpopulation with high Marker1 and Marker5 expression, while other cells (shown in yellow) are low for expression of the same markers.
```{r viz-ground-truth-celltype}
# Example
final_data_DA |>
    dplyr::filter(Sample %in% c("Baseline_S1_Set0.5", "Depletion_S1_Set0.5")) |>
    dplyr::mutate(CellType = factor(x = CellType,
                                    levels = c("target",
                                               "other"))) |>
    ggplot2::ggplot(mapping = ggplot2::aes(x = Marker1,
                                           y = Marker5,
                                           color = CellType)) +
    ggplot2::facet_grid(cols = ggplot2::vars(Sample)) +
    ggplot2::theme_bw() +
    ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5), 
                   legend.position = "none") +
    ggplot2::labs(title = "Example Marker Expression in Two Different Samples") +
    ggplot2::scale_x_continuous(transform = "log10",
                                labels = scales::trans_format(trans = "log10", 
                                                              format = scales::math_format(10^.x))) + 
    ggplot2::scale_y_continuous(transform = "log10",
                                labels = scales::trans_format(trans = "log10", 
                                                              format = scales::math_format(10^.x))) + 
    ggplot2::annotation_logticks(sides = "lb") +
    ggplot2::scale_color_viridis_d() +
    scattermore::geom_scattermore(pointsize = 3) + 
    ggh4x::force_panelsizes(cols = grid::unit(x = 2, units = "in"), 
                            rows = grid::unit(x = 2, units = "in"))
```

### Verify Ground Truth Depletion

Let's confirm the 2-fold depletion using ground truth cell label-based proportion analysis:

```{r ground-truth-proportions}
DA_res_DA <-
  final_data_DA |>
    dplyr::filter(Setting == "0.5%") |>
    dplyr::group_by(Sample, Treatment, Batch, CellType) |>
    dplyr::summarize(n = dplyr::n(), .groups = "drop") |> 
    dplyr::group_by(Sample) |>
    dplyr::mutate(prop = n / base::sum(n)) |> 
    base::split(f = ~CellType) |>
    base::lapply(X = _, FUN = function(x) {
        mod <- stats::lm(formula = log2(x = prop) ~ Treatment + Batch, data = x)
        base::c(coef = base::unname(obj = stats::coef(object = mod)["TreatmentDepletion"]),
                stats::confint(object = mod)["TreatmentDepletion",])
    }) |>
    dplyr::bind_rows(.id = "id")

ggplot2::ggplot(data = DA_res_DA, mapping = ggplot2::aes(x = coef, y = id)) +
  ggplot2::theme_bw() +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5),
                 plot.subtitle = ggplot2::element_text(hjust = 0.5)) +
  ggplot2::geom_point(size = 1) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dotted", color = "blue") +
  ggplot2::geom_vline(xintercept = -1, linetype = "dashed", color = "red", alpha = 0.5) +
  ggplot2::geom_errorbar(mapping = ggplot2::aes(xmin = `2.5 %`, xmax = `97.5 %`), width = 0.2) +
  ggplot2::labs(title = "Ground Truth Differential Abundance",
                x = "log2 Fold Change in Proportion",
                y = "CellType") +
  ggplot2::scale_y_discrete(limits = rev) +
  ggh4x::force_panelsizes(cols = grid::unit(x = 3, units = "in"), 
                          rows = grid::unit(x = 2, units = "in"))
```

Our simulation worked: we see approximately 2-fold depletion (log2FC ≈ -1) from baseline abundance.

## Differential Abundance Analysis with tinydenseR

Now let's analyze the data using `tinydenseR`, focusing on the **0.5% setting** - the most challenging scenario where the target population represents only 0.5% of cells at baseline.

### Prepare Data for tinydenseR

tinydenseR requires two main inputs:

1. **`.cells`**: A named list of file paths containing expression data per sample
2. **`.meta`**: A data frame with sample-level metadata (one row per sample)

```{r prepare-data}
# Filter to 0.5% setting
.cells.DA.0.5 <-
  final_data_DA |>
  dplyr::filter(Setting == "0.5%") |>
  dplyr::pull(var = Sample) |>
  base::unique() |>
  stats::setNames(nm = _) |>
  base::lapply(X = _, FUN = function(sample.id) {
    uri <- base::tempfile(fileext = ".RDS")
    base::saveRDS(
      object = final_data_DA[final_data_DA$Sample == sample.id,
                             c("Marker1", "Marker2", "Marker3", "Marker4", "Marker5")] |>
        base::as.matrix() |>
        base::log(),
      file = uri,
      compress = FALSE
    )
    return(uri)
  })

.meta.DA.0.5 <-
  final_data_DA |>
  dplyr::filter(Sample %in% base::names(x = .cells.DA.0.5)) |>
  dplyr::select(Sample, Treatment, Batch) |>
  dplyr::distinct() |>
  (\(x) `rownames<-`(
    x[base::match(x = base::names(x = .cells.DA.0.5), table = x$Sample), 
      c("Treatment", "Batch")],
    base::names(x = .cells.DA.0.5)
  ))()

# View metadata structure
utils::head(x = .meta.DA.0.5)
```

### Build the tinydenseR Landmark Object

The core workflow involves three main steps:

1. **`setup.lm.obj()`**: Initialize landmark object with data
2. **`get.landmarks()`**: Select landmark cells and compute PCA
3. **`get.graph()`**: Build nearest-neighbor graph and cluster landmarks
4. **`get.map()`**: Map all cells to their nearest landmarks

```{r build-landmark-object}
set.seed(seed = 123)
lm.cells.DA.0.5 <-
  tinydenseR::setup.lm.obj(
    .cells = .cells.DA.0.5,
    .meta = .meta.DA.0.5,
    .assay.type = "cyto"  # For flow/mass cytometry data
  ) |>
  tinydenseR::get.landmarks() |>
  tinydenseR::get.graph(.cl.resolution.parameter = 0.5) |>
  tinydenseR::get.map()
```

Before testing for differential abundance, let's re-examine the data structure by projecting cells from two representative samples using the PCA computed on landmark cells:

```{r viz-ground-truth-pca}
# Compute PCA projection for all cells
base::lapply(X = .cells.DA.0.5[c("Baseline_S1_Set0.5",
                                 "Depletion_S1_Set0.5")], 
             FUN = base::readRDS) |>
  base::do.call(what = rbind, args = _) |>
  (\(x) {
    ((base::t(x = x[, lm.cells.DA.0.5$pca$HVG]) - lm.cells.DA.0.5$pca$center) /
       lm.cells.DA.0.5$pca$scale) |> base::t()
  })() %*% lm.cells.DA.0.5$pca$rotation |>
  base::as.matrix() |>
  base::as.data.frame() |>
  dplyr::mutate(Sample = final_data_DA |>
                  dplyr::filter(Sample %in% base::names(x = .cells.DA.0.5[c("Baseline_S1_Set0.5",
                                                                            "Depletion_S1_Set0.5")])) |>
                  dplyr::pull(var = Sample)) |>
  ggplot2::ggplot(mapping = ggplot2::aes(x = PC1, y = PC2)) +
    ggplot2::facet_grid(cols = ggplot2::vars(Sample)) +
    ggplot2::theme_bw() +
    ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5), 
                   legend.position = "none") +
    ggplot2::labs(title = "Cell Distribution in Two Representative Samples") +
    ggplot2::stat_bin_hex(bins = 128) + 
    ggplot2::scale_fill_viridis_c(trans = "log") +
    ggh4x::force_panelsizes(cols = grid::unit(x = 2, units = "in"), 
                            rows = grid::unit(x = 2, units = "in"))
```

#### Clustering

The landmark cells are automatically clustered for visualization and traditional proportion analysis (described at the end of this vignette). Below we visualize the clustering results:

```{r viz-clustering}
tinydenseR::plotPCA(
  .lm.obj = lm.cells.DA.0.5,
  .feature = lm.cells.DA.0.5$graph$clustering$ids,,
  .plot.title = "Landmarks",
  .point.size = 2,
  .panel.size = 2
) |> 
  (\(x) x +
    ggplot2::theme(plot.subtitle = ggplot2::element_blank()) +
    ggplot2::geom_text(
      data = x$data |>
        dplyr::group_by(feature) |>
        dplyr::summarize(PC1 = stats::median(x = PC1), 
                         PC2 = stats::median(x = PC2), 
                         .groups = "drop"),
      mapping = ggplot2::aes(label = feature),
      size = 4,
      color = "black"
    )
  )()
```

#### Landmark-level visualizations

Below we visualize the landmark cells colored by batch:

```{r viz-landmarks}
tinydenseR::plotPCA(
  .lm.obj = lm.cells.DA.0.5,
  .feature = lm.cells.DA.0.5$metada$Batch[lm.cells.DA.0.5$key],
  .cat.feature.color = tinydenseR::Color.Palette[1, 1:2],
  .panel.size = 2,
  .point.size = 2,
  .color.label = "Batch",
  .plot.title = "Landmarks"
)
```

### Statistical Testing

Now we set up a design matrix and run clustering-independent statistical test:

```{r statistical-test}
# Create design matrix
.design.0.5 <-
  stats::model.matrix(object = ~ Treatment + Batch, data = .meta.DA.0.5) |> 
  (\(x) `colnames<-`(
    x = x, 
    value = base::gsub(pattern = "^Treatment|^Batch", replacement = "", 
                       x = base::colnames(x = x))
  ))()

# Run differential abundance testing
Treatment.stats.0.5 <-
  tinydenseR::get.stats(
    .lm.obj = lm.cells.DA.0.5,
    .design = .design.0.5
  )
```

### Visualizing Differential Abundance Results

#### Abundance fold changes

We can visualize the log2 fold changes in abundance at each landmark between Depletion and Baseline conditions:

```{r viz-abundance-fc}
tinydenseR::plotPCA(
  .lm.obj = lm.cells.DA.0.5,
  .feature = Treatment.stats.0.5$fit$coefficients[, "Depletion"],
  .plot.title = "Depletion vs Baseline",
  .color.label = "Abundance\nlog2(+0.5) FC",
  .panel.size = 2,
  .point.size = 2,
  .midpoint = 0
) +
  ggplot2::theme(plot.subtitle = ggplot2::element_blank())
```

#### Hypothesis testing results

We can also visualize regions of statistically significant abundance change:

```{r viz-hypothesis}
tinydenseR::plotPCA(
  .lm.obj = lm.cells.DA.0.5,
  .feature = base::ifelse(
    test = Treatment.stats.0.5$fit$coefficients[, "Depletion"] < 0,
    yes = "less abundant",
    no = "more abundant"
  ) |>
  base::ifelse(
    test = Treatment.stats.0.5$fit$pca.weighted.q[, "Depletion"] < 0.1,
    yes = _, no = "not sig."
  ) |>
  base::factor(x = _, levels = c("less abundant", "not sig.", "more abundant")),
  .plot.title = "Depletion vs Baseline",
  .color.label = "q < 0.1",
  .cat.feature.color = tinydenseR::Color.Palette[1, c(1, 6, 2)],
  .point.size = 2,
  .panel.size = 2
) +
  ggplot2::labs(subtitle = "Hypothesis Testing")
```

### Marker Expression

A scatter plot provides flexibility to explore different types of visualizations. Here we show abundance log2FC across landmarks across the distributions of Marker1 and Marker5:

```{r scatter}
tinydenseR::scatterPlot(
  .x.feature = lm.cells.DA.0.5$lm[,"Marker1"],
  .x.label = "Marker1",
  .y.feature = lm.cells.DA.0.5$lm[,"Marker5"],
  .y.label = "Marker5",
  .color.feature = Treatment.stats.0.5$fit$coefficients[, "Depletion"],
  .color.label = "Abundance\nlog2(+0.5) FC",
  .plot.title = "Depletion vs Baseline",
  .panel.size = c(3,3),
  .point.size = 2
)
```

### Beeswarm Plot

A beeswarm plot provides a comprehensive view of abundance changes across clusters:

```{r beeswarm}
tinydenseR::plotBeeswarm(
  .lm.obj = lm.cells.DA.0.5,
  .stats.obj = Treatment.stats.0.5,
  .coefs = "Depletion",
  .swarm.title = "Depletion vs Baseline",
  .row.space.scaler = 0.5,
  .perc.plot = FALSE
) +
  ggplot2::geom_vline(xintercept = -1, color = "red", linetype = "dashed")
```

Please note that the log2FC in density approaches -1 (marked by a dashed red line). The use of a pseudo-count (0.5), which is applied to handle zeros and extreme fold changes, introduces a slight bias towards underestimating differences for rare cell states.

### Cluster-based Comparisons

In addition to the high resolution analysis described above, tinydenseR can also provide cluster-based results as follows:

```{r cluster-stats}
tinydenseR::plotTradStats(.lm.obj = lm.cells.DA.0.5, .stats.obj = Treatment.stats.0.5)
```

```{r cluster-percentages, fig.height=5}
# Prepare statistical test results
stat.test.percentages.DA.0.5 <-
  lm.cells.DA.0.5$map$clustering$cell.perc |>
  dplyr::as_tibble() |>
  dplyr::mutate(treatment = lm.cells.DA.0.5$metadata$Treatment) |>
  tidyr::pivot_longer(cols = tidyr::starts_with(match = "cluster.")) |>
  dplyr::group_by(name) |>
  rstatix::t_test(formula = value ~ treatment) |>
  dplyr::mutate(
    p = Treatment.stats.0.5$trad$clustering$fit$adj.p[name, "Depletion"],
    p.adj = Treatment.stats.0.5$trad$clustering$fit$adj.p[name, "Depletion"]
  ) |>
  rstatix::add_significance() |>
  dplyr::mutate(p.adj = base::ifelse(
    test = p.adj < 0.01,
    yes = base::formatC(x = p.adj, digits = 0, format = "e"),
    no = base::formatC(x = p.adj, digits = 2, format = "f")
  )) |>
  rstatix::add_xy_position(x = "treatment")

tinydenseR::plotTradPerc(
  .lm.obj = lm.cells.DA.0.5,
  .x.split = "Treatment",
  .x.space.scaler = 0.3
) + 
  ggplot2::labs(title = "Cluster Proportions") + 
  ggpubr::stat_pvalue_manual(
    data = stat.test.percentages.DA.0.5, 
    label = "p.adj",
    label.size = 3
  ) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.05, 0.15)))
```

## Why Focus on 0.5%?

We chose to analyze the 0.5% setting in detail because:

- **Most challenging scenario**: Detecting changes in rare populations (<1%) is difficult
- **Realistic use case**: Many biologically important cell types are rare (e.g., regulatory T cells, stem cells)
- **Demonstrates sensitivity**: If tinydenseR works for 0.5%, it will work for more abundant populations
- **Clinical relevance**: Small population shifts can have significant biological impact

The ground truth analysis confirms that all three settings (0.5%, 5%, 50%) show the expected 2-fold depletion, validating our simulation design.

## Summary

In this vignette, we demonstrated how to:

1. **Simulate realistic flow cytometry data** with differential abundance and batch effects
2. **Prepare data** in the tinydenseR format (`.cells` and `.meta` objects)
3. **Build a landmark object** using the core tinydenseR workflow
4. **Run statistical tests** for differential abundance with proper sample-level replication
5. **Visualize results** using landmark-based and cluster-based approaches
6. **Interpret findings** in the context of known ground truth

### Key Advantages of tinydenseR

- **Sample-centric statistics**: Proper biological replication
- **Clustering-independent**: No arbitrary resolution parameters
- **Batch effect handling**: Natural integration into linear models
- **Scalable**: Efficient for large datasets
- **Comprehensive visualization**: Multiple complementary views of results

### Next Steps

- See the **Differential Expression Analysis** vignette for detecting state changes within populations
- See the **Trajectory Analysis** vignette for analyzing developmental or dynamic processes
- Explore additional tinydenseR features in the function documentation

## Session Information

```{r session-info}
utils::sessionInfo()
```
