---
title: "Multi-Backend Workflows with tinydenseR"
author: "Pedro Milanez-Almeida"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Multi-Backend Workflows with tinydenseR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  out.width = "100%",
  eval = FALSE
)
```

## Introduction

`tinydenseR` supports multiple single-cell data formats through a unified
S3 dispatch system. Regardless of whether your data lives in a Seurat object,
a SingleCellExperiment, an HDF5-backed AnnData file, or a bare sparse matrix,
the same analysis functions work identically.

The supported backends are:

| Backend | Input class | Result container |
|:--------|:------------|:-----------------|
| **TDRObj** (direct) | `TDRObj` | `TDRObj` |
| **Seurat** | `Seurat` | `Seurat` (TDRObj in `Misc` slot) |
| **SingleCellExperiment** | `SingleCellExperiment` | `SingleCellExperiment` (TDRObj in `metadata`) |
| **HDF5AnnData** | `HDF5AnnData` (via `anndataR`) | `TDRObj` (bare) |
| **Sparse matrix** | `dgCMatrix` | `TDRObj` (bare) |
| **On-disk matrix** | `IterableMatrix` (BPCells) | `TDRObj` (bare) |
| **Delayed matrix** | `DelayedMatrix` | `TDRObj` (bare) |
| **Cytometry** | `cytoset` (flowWorkspace) | `TDRObj` (bare) |

This means you can adopt `tinydenseR` without converting your data into a
specific format first.

## Entry Points: RunTDR()
 
`RunTDR()` is the main entry point. It accepts different input types and
runs the full pipeline (landmark selection, graph construction, optional
cell typing, mapping, and embedding). The result is returned in a form
that matches the input container when possible.

### Seurat

```{r runtdr-seurat}
library(tinydenseR)
library(SeuratObject)

seurat_result <- RunTDR(seurat_obj,
                        .sample.var = "sample_id",
                        .assay = "RNA",
                        .layer = "counts",
                        .assay.type = "RNA")
```

The returned object is the same Seurat object with the `TDRObj` stored in
`Misc(seurat_result, slot = "tdr.obj")`.

### SingleCellExperiment

```{r runtdr-sce}
library(SingleCellExperiment)

sce_result <- RunTDR(sce_obj,
                     .sample.var = "sample_id",
                     .assay = "counts",
                     .assay.type = "RNA")
```

The returned object is the same SCE with the `TDRObj` stored in
`S4Vectors::metadata(sce_result)$tdr.obj`.

### HDF5AnnData

```{r runtdr-h5ad}
library(anndataR)

adata <- read_h5ad("data.h5ad", backend = "HDF5AnnData")
h5ad_result <- RunTDR(adata,
                      .sample.var = "sample_id",
                      .assay.type = "RNA")
```

The expression matrix is converted to a BPCells on-disk format for efficient
lazy access. The result is a bare `TDRObj`.

### Sparse matrix (dgCMatrix)

```{r runtdr-dgc}
tdr <- RunTDR(dgc_matrix,
              .cell.meta = cell_metadata,
              .sample.var = "sample_id",
              .assay.type = "RNA")
```

### On-disk matrix (BPCells IterableMatrix)

```{r runtdr-bpcells}
tdr <- RunTDR(BPCells_matrix,
              .cell.meta = cell_metadata,
              .sample.var = "sample_id",
              .assay.type = "RNA")
```

For both sparse and on-disk matrix inputs, `.cell.meta` must be a
`data.frame` with one row per cell and rownames matching cell IDs.
The result is a bare `TDRObj`.

## Downstream Analysis: Identical API

Once `RunTDR()` has been called, the same pipeline code works regardless of
the input class. S3 dispatch routes each function call through the
appropriate wrapper automatically.

```{r downstream}
# These work identically whether result is Seurat, SCE, or HDF5AnnData:
result <- result |>
  get.lm(.design = design)

# Plotting works the same way
plotPCA(result)
plotBeeswarm(result, .coefs = "ConditionB")
```

This uniformity extends to all analysis and plotting functions in the
package: `get.pbDE()`, `get.specDE()`, `get.nmfDE()`,
`get.plsD()`, `plotUMAP()`, `plotHeatmap()`, and so on.

## How It Works: GetTDR / SetTDR

The dispatch mechanism follows a simple three-step pattern. When you call
an analysis function on a container object (e.g., a Seurat object), the
S3 method:

1. **Extracts** the `TDRObj` with `GetTDR()`
2. **Runs** the computation on the `TDRObj`
3. **Stores** the updated `TDRObj` back with `SetTDR()`

```
User calls:  get.lm(seurat_obj, .design = design)
                |
                v
S3 dispatch --> get.lm.Seurat(x, ...)
                |
                +-- tdr <- GetTDR(x)            # Extract TDRObj from Misc slot
                +-- tdr <- get.lm.TDRObj(tdr, ...)  # Run computation
                +-- SetTDR(x, tdr)              # Store updated TDRObj back
                |
                v
Returns:     updated seurat_obj (with results inside)
```

Where the `TDRObj` is stored depends on the container:

| Container | `GetTDR()` reads from | `SetTDR()` writes to |
|:----------|:----------------------|:---------------------|
| Seurat | `Misc(x, slot = "tdr.obj")` | `Misc(x, slot = "tdr.obj")` |
| SingleCellExperiment | `S4Vectors::metadata(x)$tdr.obj` | `S4Vectors::metadata(x)$tdr.obj` |
| HDF5AnnData | `x$uns[["tdr.obj"]]` | `x$uns[["tdr.obj"]]` |
| TDRObj | returns `x` directly | returns `tdr` directly |

## Extracting the TDRObj

You can always extract the `TDRObj` directly to inspect results or work
with them programmatically:

```{r extract-tdr}
# Extract from any container
tdr <- GetTDR(seurat_obj)

# Access linear model results
tdr$results$lm$default$fit$coefficients

# Access landmark PCA coordinates
head(tdr$landmark.embed$pca$coord)

# Access clustering assignments
tdr$landmark.annot$clustering$ids
```

The `$` accessor on a `TDRObj` maps to the underlying S4 slots, so
`tdr$results` is equivalent to `tdr@results`.

## Matrix-Class Backends

The `dgCMatrix`, `DelayedMatrix`, `IterableMatrix`, and `cytoset` backends
produce a bare `TDRObj` directly — they do not wrap the result back into a
container object. After calling `RunTDR()`, you work with the `TDRObj`
directly:
 
```{r matrix-backends}
# From a sparse matrix
tdr <- RunTDR(dgc_matrix,
              .cell.meta = cell_metadata,
              .sample.var = "sample_id",
              .assay.type = "RNA")

# All downstream functions work directly on the TDRObj
tdr <- tdr |>
  get.lm(.design = design)

plotPCA(tdr)
plotBeeswarm(tdr, .coefs = "ConditionB")
```

For `DelayedMatrix` inputs (e.g., HDF5-backed `SingleCellExperiment`
assays), the data is automatically converted to a BPCells on-disk format
for efficient lazy access before running the pipeline. You can control the
conversion directory with the `.bpcells.dir` argument.

The `cytoset` backend (from `flowWorkspace`) is designed for flow, mass,
and spectral cytometry data and requires `.assay.type = "cyto"` along
with a `.markers` argument specifying the channels to use.
