---
title: "tinydenseR Package Architecture"
author: "Pedro Milanez-Almeida"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{tinydenseR Package Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  out.width = "100%",
  eval = FALSE
)
```

## Overview

`tinydenseR` is a landmark-based framework for differential abundance and
differential expression analysis of single-cell data. It is designed for
scRNA-seq, flow, mass, and spectral cytometry data.

The core idea is to select a representative set of **landmark cells**,
build a graph over them, map all cells to their nearest landmarks, and
then model per-landmark density or expression across **samples as
biological replicates**. This preserves statistical rigor while
maintaining the richness of single-cell resolution.

## TDRObj: The Core Data Structure

All analysis results are stored in a single S4 class, `TDRObj`. It contains
12 slots that organize the data and results at each stage of the pipeline:

| Slot | Contents |
|:-----|:---------|
| `cells` | Named list of per-sample cell indices (or file paths for the files backend) |
| `metadata` | `data.frame` of sample-level metadata |
| `config` | Run parameters: key, sampling, assay.type, markers, n.threads, backend |
| `integration` | Trained projection models and batch variables (harmony.var, harmony.obj, symphony.obj, umap.model) |
| `assay` | Landmark expression layers (L × features matrices): raw, expr, scaled |
| `landmark.embed` | Landmark-space coordinate matrices (pca, le, umap), each with a `$coord` entry |
| `landmark.annot` | Per-landmark categorical annotations: clustering and celltyping, each with an `$ids` factor |
| `graphs` | Landmark-landmark connectivity: adj.matrix, snn, fgraph |
| `density` | Sample-level analytics (L × N matrices): fdens, Y, composition |
| `sample.embed` | Sample-level embeddings (N × k matrices): pca, traj, pepc |
| `cellmap` | Per-cell, per-sample data: clustering$ids, celltyping$ids, nearest.lm, fuzzy.graphs. Each entry is in-memory or an on-disk path string |
| `results` | All statistical outputs: lm, pb, marker, spec, nmf, pls, clustering, celltyping, features |

The `$` accessor is overloaded so you can use `tdr$results` instead of
`tdr@results`:

```{r accessor}
tdr$metadata              # sample-level metadata
tdr$landmark.embed$pca    # PCA coordinates of landmarks
tdr$results$lm            # linear model results
tdr$config$assay.type     # "RNA" or "cyto"
```

## Analysis Pipeline Stages

The standard analysis pipeline proceeds through a series of stages, each
populating specific slots in the `TDRObj`:

```
setup.tdr.obj() / RunTDR()
    |
    v
get.landmarks()  --> landmark.embed (pca, le)
    |                 assay (raw, expr, scaled)
    v
get.graph()      --> graphs (adj.matrix, snn)
    |                 landmark.embed (umap)
    |                 landmark.annot (clustering)
    v
get.map()        --> density (fdens, Y, composition)
    |                 cellmap (nearest.lm, fuzzy.graphs)
    v
get.lm()         --> results$lm (fit, trad)
    |                 sample.embed (pca)
    v
get.pbDE()            --> results$pb
get.pbDE(.mode="marker") --> results$marker
get.plsD()           --> results$pls
```

A typical workflow looks like:

```{r pipeline}
library(tinydenseR)

# Option 1: step-by-step
tdr <- setup.tdr.obj(...)
tdr <- get.landmarks(tdr)
tdr <- get.graph(tdr)
tdr <- get.map(tdr)
tdr <- get.lm(tdr, .design = design)

# Option 2: RunTDR() runs landmarks + graph + map + embedding in one call
result <- RunTDR(seurat_obj, .sample.var = "sample_id", .assay.type = "RNA")
result <- get.lm(result, .design = design)
```

## S3 Dispatch System

`tinydenseR` uses S3 dispatch to support multiple container types
(Seurat, SingleCellExperiment) through a single API. H5AD inputs
(either as a file path or via `anndataR`) are converted to a bare
`TDRObj` during `RunTDR()` and do not carry container dispatch
methods. The dispatch wrappers in `dispatch.R` follow one of three
patterns:

### Tier 1: Compute-Only

Functions that only need the `TDRObj` for computation. The wrapper
extracts the `TDRObj`, runs the `.TDRObj` method, and stores it back.

```{r tier1}
# Example: get.lm.Seurat
get.lm.Seurat <- function(x, ...) {
  tdr <- GetTDR(x)
  tdr <- get.lm.TDRObj(tdr, ...)
  SetTDR(x, tdr)
}
```

Functions in this tier: `get.graph()`, `get.lm()`, `get.embedding()`,
`get.plsD()`, `get.features()`,
`lm.cluster()`, `celltyping()`.

### Tier 2: Needs Source Data

Functions that need access to the original data container (e.g., to
read expression matrices via `.get_sample_matrix()`). The source object
is passed via the `.source` argument.

```{r tier2}
# Example: get.landmarks.Seurat
get.landmarks.Seurat <- function(x, ...) {
  tdr <- GetTDR(x)
  tdr <- get.landmarks.TDRObj(tdr, .source = x, ...)
  SetTDR(x, tdr)
}
```

Functions in this tier: `get.landmarks()`, `get.map()`, `get.pbDE()`.

Note: `get.markerDE()` is soft-deprecated. Use `get.pbDE(.mode = "marker")`
instead. The `get.pbDE()` function supports two modes:

- `.mode = "design"` (default) — pseudobulk differential expression
- `.mode = "marker"` — marker gene/protein identification between groups

### Special Case: `goi.summary()`

`goi.summary()` is a read-only Tier 2 function: it needs `.source` for
expression access, but it returns a summary data frame rather than
updating the container.

```{r goi}
# Example: goi.summary.Seurat
goi.summary.Seurat <- function(x, ...) {
  tdr <- GetTDR(x)
  goi.summary.TDRObj(tdr, .source = x, ...)
}
```

### Tier 3: Read-Only Plots

Plot functions only need to read the `TDRObj`; they return a plot object
rather than updating the container.

```{r tier3}
# Example: plotPCA.Seurat
plotPCA.Seurat <- function(x, ...) plotPCA.TDRObj(GetTDR(x), ...)
```

All `plot*()` functions follow this pattern: `plotPCA()`, `plotUMAP()`,
`plotBeeswarm()`, `plot2Markers()`, `plotSamplePCA()`,
`plotSampleEmbedding()`, `plotTradStats()`, `plotTradPerc()`,
`plotDensity()`, `plotPbDE()`, `plotDEA()`, `plotMarkerDE()`,
`plotHeatmap()`,
`plotPlsD()`, `plotPlsDHeatmap()`.

## Backend Data Flow

When `tinydenseR` needs to access per-sample expression matrices during
the pipeline, it calls the internal function `.get_sample_matrix()`. This
function dispatches on `config$backend` to retrieve data from the
appropriate source:

| Backend value | Source | How data is retrieved |
|:--------------|:-------|:----------------------|
| `"files"` | RDS files on disk | `readRDS(tdr@cells[[sample_idx]])` |
| `"seurat"` | Live Seurat object | `SeuratObject::LayerData(.source, assay, layer)[, col_idx]` |
| `"sce"` | Live SingleCellExperiment | `SummarizedExperiment::assay(.source, assay)[, col_idx]` |
| `"matrix"` | In-memory or on-disk matrix | Direct column indexing on the stored matrix reference |
| `"cytoset"` | flowWorkspace cytoset | `flowCore::exprs(cs[[sample_name]])` |

The `"matrix"` backend is used by `dgCMatrix`, `DelayedMatrix`,
`IterableMatrix`, and H5AD inputs (both `RunTDR.character` and
`RunTDR.HDF5AnnData`). In these cases, a reference to the matrix is
stored in a locked environment within `config$source.env` so that
`.get_sample_matrix()` can slice into it without copying the entire
matrix.

For the `"seurat"` and `"sce"` backends, the live container object is
passed as `.source` to Tier 2 functions, which then forward it to
`.get_sample_matrix()`.

## Caching System

`tinydenseR` includes an on-disk caching system for per-cell mapping
results (nearest landmark assignments, fuzzy graph memberships). This
avoids recomputing expensive cell-to-landmark mappings when re-running
downstream analyses.

The cache is managed through three user-facing functions:

```{r cache}
# Show cache location and size
tdr_cache_info(tdr)

# Validate cache integrity
tdr_cache_validate(tdr)

# Remove orphaned cache files
tdr_cache_cleanup(tdr)
```

Internal helpers (`.tdr_cache_read()`, `.tdr_cache_write()`,
`.tdr_cache_sweep_orphans()`) handle the low-level read/write operations
and are not intended to be called directly by users.
